\section{Grove Formalism}
\label{sec:Grove Formalism}

\subsection{Graphs}

$\U$ and $\W$ are separate sets of \emph{unique identifiers} (UIDs) for vertices and edges, respectively. Assume each is equipped with a total ordering $\leq$. 

$\K$ is the set of \emph{constructors} of terms in the source language, and $\P$ is the set of \emph{positions} of subterms in the source language. Each constructor $k\in\K$ is associated with a finite set of positions $\mathsf{arity}(k)\subseteq \P$, representing the set of positions that children of $k$ can inhabit.

A \emph{vertex} $v = (u, k) \in \V = \U \times \K$ is labeled with a UID, $u$, and a constructor, $k$. 

A \emph{location} $\ell = (v, p) \in \L$, represents the origin of an edge, where $v=(u,k)$ is drawn from $\V$ and position $p$ is drawn from $\mathsf{arity}(k)$.

An \emph{edge} $\e{=}(w, \ell, v) \in \E = \W \times \L \times \V$ represents a directed multi-edge 
    identified by $w$, 
    originating from location $\ell$, 
    with destination vertex $v$.

A \emph{graph} $g : \E \rightarrow \Sigma$ is a function from edges to \emph{edge states} $\Sigma = \SetOf{\bot, \Plus, \Minus}$. A graph assigns only finitely many edges to $\Plus$. The total ordering $\bot \sqsubset \Plus \sqsubset \Minus$ forms a lattice over $\Sigma$. For $s_1$, $s_2 \in \Sigma$, $s_1 \sqcup s_2$ is defined to be the least upper bound of $s_1$ and $s_2$ with respect to the $\sqsubset$ ordering. 

A graph patch command $\pi = (s,\e) \in \left\{\Plus, \Minus\right\} \times \E$ is a pair of an edge state (excluding $\bot$) and an edge. Each patch command corresponds to a transition relation between graphs:
\[
  g \overset{(s,\e)}{\longrightarrow} g\left[ \e \mapsto s \sqcup g(\e) \right]
\]

A \textit{graph patch} $\Q{\pi}$ is a sequence of graph patch commands, and the transition relation is extended to patches as the composition of the transitions of the constituent patch commands. 

\begin{lemma}[Join Semilattice]
  \label{lem:Join Semilattice}
  $(\Sigma, \sqsubset)$ with $\sqcup$ forms a join semilattice. That is, $\sqcup$ is associative, commutative, and idempotent. 
\end{lemma}

\begin{theorem}[Commutativity]
  \label{thm:Commutativity}
  For all graphs $g$ and $g'$ and all graph patch commands $\pi_1 = (s_1, \e_1)$ and  $\pi_2 = (s_2, \e_2)$, 
  if $g \overset{\pi_1 \pi_2}{\longrightarrow} g'$,
  then $g \overset{\pi_2 \pi_1}{\longrightarrow} g'$
\end{theorem}

The commutativity of patch commands generalizes directly to the commutativity of patches. 

\subsection{Groves}

\figureTermSyntaxContent

A live subgraph $G$, which is a finite set of edges, decomposes into a grove $\Theta$, which is a finite set of terms.


\begin{align*}
    \mathsf{parents}_{G}(v) &= \SetOf{v' \in \V \SuchThat{(w, (v', p), v) \in G \land w \in \W \land p \in \P}}\\
    \mathsf{children}_{G}(\ell) &= \SetOf{(w,v) \in \W\times\V \SuchThat{(w, \ell, v) \in G}}
\end{align*}


A vertex is an \textit{NP root} in $G$ if $\mathsf{parents}_{G}(v)=\emptyset$.\\

A vertex is an \textit{MP root} in $G$ if $\abs{\mathsf{parents}_{G}(v)} > 1$.\\

A vertex is a \textit{U root} in $G$ if $\exists \{v_i\}_{1 \leq i \leq n}$ such that $v_1 = v_n = v$, $\mathsf{parents}_{G}(v_i)=\{v_{i+1}\}$ for all $1 \leq i < n$, and the UID of $v$ is the minimum of the UIDs of $\{v_i\}_{1 \leq i \leq n}$.\\ 

$\mathsf{root}_G(v)$ if $v$ is an NP root, an MP root, or a U root in $G$.


\begin{lemma}
\label{lem:uniq-classification}
    (Unique Classification) For any live subgraph $G\subseteq \E$ and vertex $v\in V$, exactly one of the following holds: 
    \begin{itemize}
        \item $v$ is an NP root in $G$
        \item $v$ is an MP root in $G$
        \item $v$ is an U root in $G$
        \item $v$ is not a root in $G$
    \end{itemize}
\end{lemma}


    $\mathsf{decompVertex}_{G}(v = (u,k)) = \genericTerm{u}{\mathsf{decompLoc}_{G}((v,p))}$



\begin{align*}
    \mathsf{decompLoc}_{G}(\ell) &= \begin{cases}
        \ehole &\quad \mathsf{children}_{G}(\ell) = \emptyset \\ \eid{w}{}{\mathsf{decompChild}_G(w, v')} &\quad \mathsf{children}_{G}(\ell) = \SetOf{(w,v')}\\ \conflictHoleTerm{\ell}{\eid{w}{}{\mathsf{decompChild}_G(w, v')} \SuchThat{(w,v') \in \mathsf{children}_{G}(\ell)}} &\quad \abs{\mathsf{children}_{G}(\ell)} > 1 \\
    \end{cases}
\end{align*}



\begin{align*}
    \mathsf{decompChild}_G(w, v)  &= \begin{cases}
        \multiref  &\quad v\text{ is an MP root in } G\\
        \uniref  &\quad v\text{ is a U root in} G\\
        \mathsf{decompVertex}_{G}(v) &\quad v\text{ is not a root in} G
    \end{cases} 
\end{align*}



    $\mathsf{vertices}(G) = \SetOf{v \in \V \SuchThat{(w, (v,p), v') \in G \land w \in \W \land v' \in \V \land p \in \P}}$



    $\mathsf{decomp}(G) = \SetOf{\mathsf{decompVertex}_{G}(v) \SuchThat{\mathsf{root}_G(v)} \land v \in \mathsf{vertices}(G)}$



    \begin{align*}
        \mathsf{recompTerm}(\genericTerm{u}{\subtermMV_p}) &= \bigcup_{p\in \mathsf{arity}(k)}\mathsf{recompChildTerm}(((u,k),p), \subtermMV_p)\\
        \mathsf{recompTerm}(\multiref) &= \emptyset\\
        \mathsf{recompTerm}(\uniref) &= \emptyset
    \end{align*}



    \begin{align*}
    \mathsf{recompChildTerm}(\ell, \ehole) &= \emptyset\\
    \mathsf{recompChildTerm}(\ell, \eid{w}{}\termMV) &= \SetOf{(w, \ell, \mathsf{vertexOfTerm}(\termMV))}\cup \mathsf{recompTerm}(t)\\
    \mathsf{recompChildTerm}(\ell, \conflictHoleTerm{\ell}{\eid{w_i}{}\termMV_i}_{i<n}) &= \bigcup_{i<n} (\SetOf{(w_i, \ell, \mathsf{vertexOfTerm}(\termMV_i))}\cup \mathsf{recompTerm}(t_i))
    \end{align*}
    


    \begin{align*}
    \mathsf{vertexOfTerm}(\genericTerm{u}{\subtermMV_p}) &= (u,k)\\
    \mathsf{vertexOfTerm}(\multiref) &= v\\
    \mathsf{vertexOfTerm}(\uniref) &= v
    \end{align*}
    


    $\mathsf{recomp}(T) = \bigup{\termMV \in T}\mathsf{recompTerm}(\termMV)$


\begin{theorem}
\label{thm:Recomposability}
    (Recomposability) For all subgraphs $G$, $\mathsf{recomp}(\mathsf{decomp}(G))=G$.
\end{theorem}

\subsection{Partitioned Groves}

A partitioned grove $\gamma$ is a quadruple $(\subtermMV_r, NP, MP, U)$ where $\subtermMV_r \in ChildTerm$ and $NP, MP, U$ are finite sets of terms. A grove $\mathsf{decomp}(G)$ is partitioned with respect to a \textit{distinguished root} location $\ell_r=(v_r=(u_r,k_r),p_r)$ such that $v_r$ is an NP root in $G$ and $\mathsf{arity}(k_r)=\SetOf{p_r}$.

\begin{align*}
    \mathsf{distinguishedChildTerm}(\ell_r=((u_r,k_r),p_r), \Theta) &= \subtermMV_r &\quad \text{ if }
    \eid{u_r}{k_r}\ {\subtermMV_r}\in \Theta \\
    \mathsf{distinguishedChildTerm}(\ell_r=((u_r,k_r),p_r), \Theta) &= \emptyHole{\ell_r} &\quad \text{ otherwise}
\end{align*}



    \begin{align*}
    \mathsf{rootVertexOfTerm}_{G}(\genericTerm{u}{\subtermMV_p}) &= (u,k)\\
    \mathsf{rootVertexOfTerm}_{G}(\multiref) &= v'\mathsf{ where }(w, (v',p), v)\in G\\
    \mathsf{rootVertexOfTerm}_{G}(\uniref) &=  v'\mathsf{ where }(w, (v',p), v)\in G
    \end{align*}
    


    \begin{align*}
    \mathsf{grove}(\ell_r=(v_r,p_r), \Theta, G) = (&\mathsf{distinguishedChildTerm}(\ell_r, \Theta),\\ &\SetOf{t\in \Theta \SuchThat{ \mathsf{rootVertexOfTerm}_{G}(t)\neq v_r\text{ and is an NP root in }G}},\\
    &\SetOf{t\in \Theta \SuchThat{ \mathsf{rootVertexOfTerm}_{G}(t)\text{ is an MP root in }G}},\\ 
    &\SetOf{t\in \Theta \SuchThat{ \mathsf{rootVertexOfTerm}_{G}(t)\text{ is a U root in }G}}) 
    \end{align*}

\subsection{User Edit Actions}

\figureUserActionSyntax


\label{def:edges}
    \[
    \begin{array}{ccc}
    \mathsf{edges}_G(\ehole) & = & \emptyset\\
    \mathsf{edges}_G(\conflict{\termMV}) & = & \SetOf{(w_i,\ell,v)\SuchThat{(w_i,\ell,v)\in G\land i<n\land \ell\in\L\land v\in\V}}\\
    \mathsf{edges}_G(\genericTerm{u}{\subtermMV_p}) & = & \SetOf{(w,\ell,(u,k))\SuchThat{(w,\ell,(u,k))\in G\land w\in\W\land \ell\in\L}}\\
    \mathsf{edges}_G(\multiref) & = & \SetOf{(w,\ell,v)\SuchThat{(w,\ell,v)\in G\land\ell\in\L}}\\
    \mathsf{edges}_G(\uniref) & = & \SetOf{(w,\ell,v)\SuchThat{(w,\ell,v)\in G\land\ell\in\L}}\\
    \end{array}
\]


Each $k\in\K$ in the language is equipped with a designated position $\mathsf{defaultPos}(k)\in \mathsf{arity}(k)$. 

\figureUserActionTransitionsContent

